DNA Methylation Hierarchical Variance Model\nStan Results

### 2021-01-06 StanAnalysis.Rmd
Edit to StanAnalysis

\subsection{Prior Probabilities}
Priors for each parameter (mu, betaT, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution
\newline Additionally: Prior relaxation was performed for the Cauchy and gamma distribtions by algebraically increasing the variance of the distributions while maintaining the same mode (peak)

\subsection{Conservation score}
We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e. lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP). Therefore we choose the log-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation.
We take the median of the posterior distribution for this parameter. Additionally, we calculate a regularization term based on the inverse of the L2 norm of sigmaP and sigmaT.

<!--  $$ \text{score}_\text{prob} = \Pr (\text{log}\frac{\sigma_{P}}{\sigma_T} > 0) = \int \text{log}\frac{\sigma_{P}}{\sigma_T} > 0 ~ dp $$
$$ \text{score}_\text{mean} = \text{Mean} (\text{log}\frac{\sigma_{P}}{\sigma_T}) ,    ~ -->
$$\text{score}_\text{med} = \text{Median} (\text{log}\frac{\sigma_{P}}{\sigma_T}) $$

@@

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu") + theme(aspect.ratio = 1)
pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "BetaT") + theme(aspect.ratio = 1)
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE") + theme(aspect.ratio = 1)
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP") + theme(aspect.ratio = 1)
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT") + theme(aspect.ratio = 1)
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT") + theme(aspect.ratio = 1)
# log sigmas
#pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.34*dgamma(exp(x), 2.7693, 7.0029), 0.25, c(-5,5), "log(SigmaP)")
#pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.42*dgamma(exp(x), 2.1457, 2.4819), 0.25, c(-5,5), "log(SigmaPT)")
#pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.22*dgamma(exp(x), 3.3643, 10.430), 0.25, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))

# arrange each parameter as a subplot # ,labels=c("M","B","E","P","PT","T","lP","lPT","lT")
p_all <- ggarrange(pM,pB,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3) #,pLP,pLPT,pLT
print(p_all)
# ggsave("posteriors_geneassoc.svg",p_all)
rm(pM,pB,pE,pP,pPT,pT)
```

@@

# plot the two conservations score prob sums
pP1 <- post_histogram(CpGscore_full$median,genemask,0.1,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")")) + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
# ggsave("CpGscore_geneassoc.svg")
# prob1_proxy = log(sigmaP_full$p50/sigmaT_full$p50)
# pP2 <- post_histogram(prob1_proxy,genemask,0.05,c(-3,3),expression(" log( median" ~sigma[P]/~"median"~sigma[T]~")"))
# # regularized version
# L2norm = sqrt(sigmaP_full$p50^2 + sigmaT_full$p50^2)
# prob1_reg = log(sigmaP_full$p50/sigmaT_full$p50) - 1/L2norm
# pP3 <- post_histogram(prob1_reg,genemask,0.1,c(-10,5),"Regularized")
# p_probs <- ggarrange(pP1,pP2,pP3,nrow=1,ncol=3)
# print(p_probs)
```

\newpage
\subsection{Summary of neff and Rhat for each parameter}
```{r neffRhat, echo=FALSE, warning=FALSE}
# for each parameter (mu, betaT, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
# reporting: mean n_eff, min n_eff, # n_eff<500, %, # n_eff<100, %
n_eff_mat <-
  data.frame(
  mu = mu_full$n_eff,
  betaT = betaT_full$n_eff,
  sigmaP = sigmaP_full$n_eff,
  sigmaPT = sigmaPT_full$n_eff,
  sigmaT = sigmaT_full$n_eff,
  sigmaE = sigmaE_full$n_eff,
  lp = lp_full$n_eff
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  min = min(value),
  s500 = sum(value < 500),
  p500 = 100 * mean(value < 500),
  s100 = sum(value < 100),
  p100 = 100 * mean(value < 100),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","betaT","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

@@

med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, betaT=betaT_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
  mutate(score_median=CpGscore_full$median)

@@

cat("Permutate data with bootstrapping")
ptm<-proc.time()
# 2nd function of methcon5
permed_data <- gene_summary %>%
  filter(n < quantile(n, 0.99)) %>%
  boot_naive(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
  boot_naive(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaP, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaPT, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, sigmaT, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000)
write_csv(permed_data, here(prepdatafolder, "permed_data.csv"))
cat(proc.time()-ptm)

@@

``` {r enrich, include=FALSE}
# expression data
# downloaded file: E-MTAB-2836-query-results.tpms.tsv
# source: https://www.ebi.ac.uk/gxa/experiments/E-MTAB-2836/Results
expression_atlas <- read_tsv(here("data-raw", "E-MTAB-2836-query-results.tpms.tsv"), skip = 4, 
                             col_types = cols(
                               .default = col_double(),
                               `Gene ID` = col_character(),
                               `Gene Name` = col_character()
                             ))

# Gene Set Enrichment
tidy_gsea <- function(genes) {
  AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, genes, 'ENTREZID', 'SYMBOL') %>%
    ReactomePA::enrichPathway()
}

enrich_sigmaP <- permed_data %>%
  #filter(sigmaP_v3 < quantile(sigmaP_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaP_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaPT <- permed_data %>%
  #filter(sigmaPT_v3 < quantile(sigmaPT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaPT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_sigmaT <- permed_data %>%
  #filter(sigmaT_v3 < quantile(sigmaT_v3, 0.05, na.rm = TRUE)) %>%
  filter(sigmaT_v3 < 0.05) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
enrich_score_median <- permed_data %>%
  #filter(score_median_v3 > quantile(score_median_v3, 0.95, na.rm = TRUE)) %>%
  filter(score_median_v3 > 0.95) %>%
  pull(UCSC_RefGene_Name) %>% tidy_gsea()
```

@@

# plot: n_sites, mu, betaT, sigmaP, sigmaPT, sigmaT
n_cutlo = 5
n_cuthi = quantile(gene_summary$n, 0.99)
pN <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
geom_vline(xintercept=n_cuthi,linetype = "dashed",color="red") +annotate(geom="text",x=120,y=0.013,size=2,label=sprintf("# genes >%0.1f sites:\n %i / %i (99th pctile)", n_cuthi,sum(gene_summary$n>n_cuthi),nrow(gene_summary)))
gene_summary <- gene_summary %>% filter(n>=n_cutlo) %>% filter(n<n_cuthi)
pM <- post_histogram(gene_summary$mu,0.1,c(-5,5),"Mu")
pB <- post_histogram(gene_summary$betaT,0.1,c(-5,5),"BetaT")
pP <- post_histogram(gene_summary$sigmaP,0.05,c(0,2),"SigmaP")
pPT <- post_histogram(gene_summary$sigmaPT,0.05,c(0,2),"SigmaPT")
pT <- post_histogram(gene_summary$sigmaT,0.05,c(0,2),"SigmaT")
pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")

# arrange each parameter as a subplot
p_all <- ggarrange(pN,pM,pB,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pN,pM,pB,pP,pPT,pT,pP1)

@@

\newpage
\subsection{Gene-average parameters vs N sites}
```{r Params_vsN, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA, dev = "png"}
pM <- ggplot(gene_summary) + geom_point(aes(x=n,y=mu), alpha=0.01) + ggtitle("Mu vs N-sites")
pB <- ggplot(gene_summary) + geom_point(aes(x=n,y=betaT), alpha=0.01) + ggtitle("BetaT vs N-sites")
pE <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaE), alpha=0.01) + ggtitle("SigmaE vs N-sites")
pP <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaP), alpha=0.01) + ggtitle("SigmaP vs N-sites")
pPT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaPT), alpha=0.01) + ggtitle("SigmaPT vs N-sites")
pT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaT), alpha=0.01) + ggtitle("SigmaT vs N-sites")

@@

\newpage
\subsection{Figure 3: Reactome Pathway Enrichment Analysis}
Enrichment defined from the group of genes <5th percentile of each sigma, or >95th percentile of conservation score.
```{r Fig3, echo=FALSE, fig.height=3.5, comment=NA}
cat("Number of enrichment pathways (sigmas):")
c(sigmaP = length(enrich_sigmaP$Description),
  sigmaPT = length(enrich_sigmaPT$Description),
  sigmaT = length(enrich_sigmaT$Description)) %>%
  sort()

list(
  sigmaP = enrich_sigmaP$Description,
  sigmaPT = enrich_sigmaPT$Description,
  sigmaT = enrich_sigmaT$Description
) %>%
  UpSetR::fromList() %>%
  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c(3, 2, 1)])
grid::grid.text("Figure 3A: Frequency of overlap between pathways that are called \nas most conserved for each hierarchical variance.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))

cat("Number of enrichment pathways (probs):")
c(score_median = length(enrich_score_median$Description),
  score_median2 = length(enrich_score_median2$Description),
  score_reg = length(enrich_score_reg$Description)) %>%
  sort()

list(
  score_median = enrich_score_median$Description,
  score_median2 = enrich_score_median2$Description,
  score_reg = enrich_score_reg$Description
) %>%
  UpSetR::fromList() %>%
  UpSetR::upset(sets.bar.color = RColorBrewer::brewer.pal(3, "Set2")[c( 2, 1)])
grid::grid.text("Figure 3B: Frequency of overlap between pathways that are called \nas most conserved for each conservation score.", x=0.65, y=0.05, gp=grid::gpar(fontsize=10))
```

\newpage
\subsection{Figure 4: Pathway Enrichment Network}
```{r Fig4, echo=FALSE, message=FALSE, fig.height=2.5}
# Figure 4
cat("sigmaP:")
ggg <- enrich_sigmaP %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4A: Relationship between pathways that are called as conserved based on sigmaP.", title = NULL)

ggg$layers <- ggg$layers[1:2]

ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

#ggsave(filename = here("figures", "figure-4.png"), width = 7, dpi = 320, height = 4.326)

# sigmaPT
cat("sigmaPT:")
ggg <- enrich_sigmaPT %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4B: Relationship between pathways that are called as conserved based on sigmaPT.", title = NULL)

ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

# sigmaT
cat("sigmaT:")
ggg <- enrich_sigmaT %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4C: Relationship between pathways that are called as conserved based on sigmaT.", title = NULL)

ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")
```

\newpage
```{r Fig4DEF, echo=FALSE, message=FALSE, fig.height=2.5}
# score_median
cat("score_median")
ggg <- enrich_score_median %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4D: Relationship between pathways that are called as conserved based on score_median.", title = NULL)
ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

# score_median2
cat("score_median2")
ggg <- enrich_score_median2 %>%
  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
  labs(caption = "Figure 4E: Relationship between pathways that are called as conserved based on score_median2.", title = NULL)
ggg$layers <- ggg$layers[1:2]
ggg + coord_flip() + 
  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
  scale_color_gradientn(colors = "black") +
  guides(color = "none", size = "none")

# score_reg
cat("score_reg")
#ggg <- enrich_score_reg %>%
#  ReactomePA::emapplot(showCategory = 30, layout = "graphopt") +
#  labs(caption = "Figure 4G: Relationship between pathways that are called as conserved based on score_reg.", title = NULL)
#ggg$layers <- ggg$layers[1:2]
#ggg + coord_flip() +
#  ggraph::geom_node_label(aes_(label = ~name), repel = TRUE, size = 2) +
#  scale_color_gradientn(colors = "black") +
#  guides(color = "none", size = "none")
```

\subsection{Figure 5: Gene Conservation vs Gene Expression}
```{r Fig5, echo=FALSE, message=FALSE}
# Figure 5A
figure5_data <- permed_data %>% 
  left_join(by = c("UCSC_RefGene_Name" = "gene"), promoted_data) %>%
  left_join(expression_atlas %>% select(`Gene Name`, colon),
            by = c("UCSC_RefGene_Name" = "Gene Name"))

bind_rows(
  figure5_data %>%
    select(sigmaP_v3, sigmaP_pro, colon) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(sigmaPT_v3, sigmaPT_pro, colon) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(sigmaT_v3, sigmaT_pro, colon) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon)
) %>%
  group_by(name) %>%
  mutate(value = factor(11-as.numeric(cut_number(value, 10)))) %>%
  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
  mutate(type = factor(type, c("sigmaP","sigmaPT","sigmaT"), c("SigmaP-Colon","SigmaPT-Colon","SigmaT-colon")),
         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
  ggplot(aes(value, expression, fill = type)) +
  geom_boxplot() +
  scale_y_log10() +
  guides(color = "none") +
  facet_grid(method ~ type) +
  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
       caption = "Figure 5A: Relationship between hierarchical variance (Sigmas) and gene expression.") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")

# Figure 5B
bind_rows(
  figure5_data %>%
    select(score_median_v3, score_median_pro, colon) %>%
    rename(scoremedian_v3 = score_median_v3, scoremedian_pro = score_median_pro) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(score_median2_v3, score_median2_pro, colon) %>%
    rename(scoremedian2_v3 = score_median2_v3, scoremedian2_pro = score_median2_pro) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon),
  figure5_data %>%
    select(score_reg_v3, score_reg_pro, colon) %>%
    rename(scorereg_v3 = score_reg_v3, scorereg_pro = score_reg_pro) %>%
    pivot_longer(-colon) %>%
    drop_na() %>%
    rename(expression = colon)
) %>%
  group_by(name) %>%
  mutate(value = factor(as.numeric(cut_number(value, 10)))) %>%
  separate(name, c("type", "method"), sep = "_", extra = "merge") %>%
  mutate(type = factor(type, c("scoremedian","scoremedian2","scorereg"), c("ScoreMedian-colon","ScoreMedian2-colon","Scorereg-colon")),
         method = factor(method, c("v3", "pro"), c("Boot-strapped values", "Promoter region"))) %>%
  ggplot(aes(value, expression, fill = type)) +
  geom_boxplot() +
  scale_y_log10() +
  guides(color = "none") +
  facet_grid(method ~ type) +
  labs(x = "quantile of conservation by gene", y = "Expression", title = NULL,
       caption = "Figure 5B: Relationship between conservation scores and gene expression.") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

\newpage
\section{Appendix: Enrichment Gene/Pathway Lists}
\subsection{A1: SigmaP}
Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
```{r App_sigmaP, echo=FALSE, fig.fullwidth=TRUE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>%
  select(UCSC_RefGene_Name, sigmaP, sigmaP_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaP_v3) %>%
  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_sigmaP$Description, ngenes = enrich_sigmaP$Count, padj = enrich_sigmaP$p.adjust) %>% arrange(padj)
geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
  lapply(function(x) paste(x,collapse=";")) %>% unlist
enrichedPaths$Genes = geneList

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_sigmaP$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 30
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

library(pander)
enrichedGenes %>% pander()

cat("SigmaP top 30 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A2: SigmaPT}
Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
```{r App_sigmaPT, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>%
  select(UCSC_RefGene_Name, sigmaPT, sigmaPT_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaPT_v3) %>%
  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_sigmaPT$Description, ngenes = enrich_sigmaPT$Count, padj = enrich_sigmaPT$p.adjust) %>% arrange(padj)
geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
  lapply(function(x) paste(x,collapse=";")) %>% unlist
enrichedPaths$Genes = geneList

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_sigmaPT$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 30
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

enrichedGenes %>% pander()

cat("\nSigmaPT top 30 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A3: SigmaT}
Genes are selected with cutoff <5th percentile (adjusted bootstrap), showing top 30 genes/pathways
```{r App_sigmaT, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>%
  select(UCSC_RefGene_Name, sigmaT, sigmaT_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Pctile=sigmaT_v3) %>%
  filter(Pctile < quantile(Pctile, 0.05, na.rm = TRUE)) %>%
  arrange(Pctile)

enrichedPaths <- tibble(Pathway = enrich_sigmaT$Description, ngenes = enrich_sigmaT$Count, padj = enrich_sigmaT$p.adjust) %>% arrange(padj)
geneList <- enrich_reactome$geneID %>% strsplit(split='/') %>%
  lapply(function(x) AnnotationDbi::mapIds(org.Hs.eg.db::org.Hs.eg.db, x, 'SYMBOL', 'ENTREZID')) %>%
  lapply(function(x) paste(x,collapse=";")) %>% unlist
enrichedPaths$Genes = geneList

ngene <- dim(enrichedGenes)[1]
npath <- length(enrich_sigmaT$Description)

cat(paste0("Total number of genes: ", ngene, "     Number of enrichment pathways: ", npath))

nshow = 30
if (ngene > nshow) {
  enrichedGenes <- enrichedGenes[1:nshow,]
}
if (npath > nshow) {
  enrichedPaths <- enrichedPaths[1:nshow,]
}

enrichedGenes %>% pander()

cat("SigmaT top 30 enriched paths:  \n")
enrichedPaths %>% pander()
```

\newpage
\subsection{A6: Score Median}
Genes are selected with cutoff >95th percentile (adjusted bootstrap), showing top 50 genes/pathways
```{r App_score_median, echo=FALSE, comment=NA, message=FALSE}
enrichedGenes <- permed_data %>% 
  select(UCSC_RefGene_Name, score_median, score_median_v3, n) %>%
  rename(Gene=UCSC_RefGene_Name, Score_median=score_median, Pctile=score_median_v3) %>%
  filter(Pctile > quantile(Pctile, 0.95, na.rm = TRUE)) %>%
  arrange(Pctile)

>>>>

\subsection{Prior Probabilities}
Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (nu), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution
\newline Additionally: Prior relaxation was performed for the Cauchy and gamma distribtions by algebraically increasing the variance of the distributions while maintaining the same mode (peak)

\subsection{Conservation score}
We want to understand which sites of DNA methylation are fundamentally conserved within tumors, i.e. lower variation in methlyation within tumor (sigmaT) relative to the variation within normal healthy tissue (sigmaP). Therefore we choose the log2-ratio of sigmaP/sigmaT to represent a conservation score, which is positive when tumor variation is lower than normal variation.
We take the median of the posterior distribution for this parameter.

$$\text{score}_\text{med} = \text{Median} (\text{log}_2\frac{\sigma_{P}}{\sigma_T}) $$

@@

# plot mu,nu,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.3*fBG(x,0.3557,-3.0675, 1.3027, 1.1290, 2.4110), 0.25, c(-5,5), "Mu") + theme(aspect.ratio = 1)
pN <- post_prior_histogram(nu_full$p50, genemask, function(x) 1.5*dcauchy(x,-0.04051, 0.5799), 0.25, c(-5,5), "Nu") + theme(aspect.ratio = 1)
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 3.3*dgamma(x,2,2), 0.05, c(0,2), "SigmaE") + theme(aspect.ratio = 1)
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 1.52*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP") + theme(aspect.ratio = 1)
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.6*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT") + theme(aspect.ratio = 1)
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 0.97*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT") + theme(aspect.ratio = 1)
# log sigmas
#pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.34*dgamma(exp(x), 2.7693, 7.0029), 0.25, c(-5,5), "log(SigmaP)")
#pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.42*dgamma(exp(x), 2.1457, 2.4819), 0.25, c(-5,5), "log(SigmaPT)")
#pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.22*dgamma(exp(x), 3.3643, 10.430), 0.25, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6))

# arrange each parameter as a subplot # ,labels=c("M","N","E","P","PT","T","lP","lPT","lT")
p_all <- ggarrange(pM,pN,pE,pP,pPT,pT,hjust=0,nrow=2,ncol=3) #,pLP,pLPT,pLT
print(p_all)
# ggsave("posteriors_geneassoc.svg",p_all)
rm(pM,pN,pE,pP,pPT,pT)
```

@@

# plot the two conservations score prob sums
pP1 <- post_histogram(CpGscore_full$median,genemask,0.1,c(-3,3),expression("Median log("~sigma[P]/sigma[T]~")")) + scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
# ggsave("CpGscore_geneassoc.svg")
```

\newpage
\subsection{Summary of neff and Rhat for each parameter}
```{r neffRhat, echo=FALSE, warning=FALSE}
# for each parameter (mu, nu, sigmaP, sigmaPT, sigmaT, sigmaE, lp__)
# reporting: mean n_eff, min n_eff, # n_eff<500, %, # n_eff<100, %
n_eff_mat <-
  data.frame(
  mu = mu_full$n_eff,
  nu = nu_full$n_eff,
  sigmaP = sigmaP_full$n_eff,
  sigmaPT = sigmaPT_full$n_eff,
  sigmaT = sigmaT_full$n_eff,
  sigmaE = sigmaE_full$n_eff,
  lp = lp_full$n_eff
  ) %>% gather(param,value,mu:lp) %>% group_by(param) %>% summarize(
  mean = mean(value),
  min = min(value),
  s500 = sum(value < 500),
  p500 = 100 * mean(value < 500),
  s100 = sum(value < 100),
  p100 = 100 * mean(value < 100),
  .groups = "drop"
  ) %>% arrange(match(.$param,c("mu","nu","sigmaP","sigmaPT","sigmaT","sigmaE","lp")))

@@

med_vals <- tibble(Name=FullAnnotation[,1], mu=mu_full$p50, nu=nu_full$p50, sigmaE=sigmaE_full$p50, sigmaP=sigmaP_full$p50, sigmaPT=sigmaPT_full$p50, sigmaT=sigmaT_full$p50, Rhat=lp_full$Rhat) %>% mutate(logP=log(sigmaP), logPT=log(sigmaPT), logT=log(sigmaT)) %>%
  mutate(score_median=CpGscore_full$median)

@@

cat("Permutate data with bootstrapping")
ptm<-proc.time()
# 2nd function of methcon5
permed_data <- gene_summary %>%
  filter(n < quantile(n, 0.99)) %>%
  boot_naive(gened_data, score_median, UCSC_RefGene_Name, 1000) %>%
  boot_adj(gened_data, score_median, UCSC_RefGene_Name, 1000)
write_csv(permed_data, here(prepdatafolder, "permed_data.csv"))
cat(proc.time()-ptm)

@@

@@

# plot: n_sites, mu, nu, sigmaP, sigmaPT, sigmaT
n_cutlo = 5
n_cuthi = quantile(gene_summary$n, 0.99)
pNS <- post_histogram(gene_summary$n,1,c(0,200),paste0("N sites (max = ",max(gene_summary$n),")")) + geom_vline(xintercept=n_cutlo-0.5,linetype = "dashed",color="red") + annotate(geom="text",x=80,y=0.03,size=2,label=paste0("# genes <",n_cutlo," sites:\n", sum(gene_summary$n<n_cutlo)," / ",nrow(gene_summary))) +
geom_vline(xintercept=n_cuthi,linetype = "dashed",color="red") +annotate(geom="text",x=120,y=0.013,size=2,label=sprintf("# genes >%0.1f sites:\n %i / %i (99th pctile)", n_cuthi,sum(gene_summary$n>n_cuthi),nrow(gene_summary)))
gene_summary <- gene_summary %>% filter(n>=n_cutlo) %>% filter(n<n_cuthi)
pM <- post_histogram(gene_summary$mu,0.1,c(-5,5),"Mu")
pN <- post_histogram(gene_summary$nu,0.1,c(-5,5),"Nu")
pP <- post_histogram(gene_summary$sigmaP,0.05,c(0,2),"SigmaP")
pPT <- post_histogram(gene_summary$sigmaPT,0.05,c(0,2),"SigmaPT")
pT <- post_histogram(gene_summary$sigmaT,0.05,c(0,2),"SigmaT")
pP1 <- post_histogram(gene_summary$score_median,0.05,c(-3,3),"Score_median")

# arrange each parameter as a subplot
p_all <- ggarrange(pNS,pM,pN,pP,pPT,pT,pP1,hjust=0,nrow=3,ncol=3)
print(p_all)
rm(pNS,pM,pN,pP,pPT,pT,pP1)

@@

\newpage
\subsection{Gene-average parameters vs N sites}
```{r Params_vsN, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA, dev = "png"}
pM <- ggplot(gene_summary) + geom_point(aes(x=n,y=mu), alpha=0.01) + ggtitle("Mu vs N-sites")
pB <- ggplot(gene_summary) + geom_point(aes(x=n,y=nu), alpha=0.01) + ggtitle("Nu vs N-sites")
pE <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaE), alpha=0.01) + ggtitle("SigmaE vs N-sites")
pP <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaP), alpha=0.01) + ggtitle("SigmaP vs N-sites")
pPT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaPT), alpha=0.01) + ggtitle("SigmaPT vs N-sites")
pT <- ggplot(gene_summary) + geom_point(aes(x=n,y=sigmaT), alpha=0.01) + ggtitle("SigmaT vs N-sites")

@@
